Title here

### 2020-09-01 analysis/paper/paper.Rmd
casi draft

---
twocol: TRUE  #  Use FALSE for submission. 
title: Title here
author1: 
  name: Elio Campitelli
  email: elio.campitelli@cima.fcen.uba.ar
author2: 
  name: Leandro Díaz
affiliation: "CIMA UBA blablabla"
extra_authors: 
  - name: Carolina Vera
abstract: |
  Enter the text of your abstract here. This is a sample American Meteorological Society (AMS) \LaTeX\ template.  This document provides authors with instructions on the use of the AMS \LaTeX\ template.  Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document. The template.tex file should be modified by authors for their own manuscript.
statement: |
  This is significant becasue I wrote it.
bibliography: AsymSAM.bib
output: 
  rticles::ams_article:
    latex_engine: "xelatex"
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
---

```{r setup, echo = FALSE,warning=FALSE, message = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.path = "",
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.extra = 42,
  cache.path = "../cache/",
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  out.extra = ""
)
library(metR)
library(data.table)
library(ggplot2)
library(magrittr)
library(ggperiodic)
library(asymsam)
library(latex2exp)
library(tagger)

theme_set(theme_asymsam(base_size = 10))
ZeroBreaks <- AnchorBreaks(0, NULL, 0)
axis_labs_smol <- theme(axis.text = element_text(size = 6))
lev.lab <- AddSuffix(" hPa")

@@

combine_lists <- function(...) {
  X <- list(...)
  names <- names(X[[1]])
  names(names) <- names
  lapply(names, function(n) {
    unlist(lapply(X, function(x) x[[n]]))
  })
}
```

@@

```{r compute-SAM}
SAM <- hgt[lat <= -20] %>% 
  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
  .[, .(eof = list(eof_flip(EOF(hgt ~ time | lon + lat, n = 1, data = .SD)))), 
    by = .(lev)]

@@

We used monthly geopotential height at 2.5 longitude by 2.5 latitude resolution from ERA5 [@hersbach] for the period 1979 to 2018 (inclusive).

@@

```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height. Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right). Arbitrary units.", fig.width=6, fig.height=2.5, fig.env="figure*"}
sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 
  setnames("hgt", "full") %>% 
  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev)] %>% 
  rm_singleton()

lab_sam <-  c(full = "Full", 
              asym = "Asymmetric",
              sym  = "Symmetric")

@@

sam_plot <- sam_sep %>% 
  .[lev %in% c(700)] %>%
  melt(id.vars = c("lon", "lat", "lev")) %>% 
  .[variable == "sym", value := value + rnorm(.N, 0, sd = 1e-6)] %>%
  .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)]

@@

sam_plot %>% 
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +
  # geom_contour(aes(z = value, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
  geom_contour_fill(aes(z = value), global.breaks = FALSE, breaks = ZeroBreaks) +
  geom_contour_tanaka(aes(z = value), global.breaks = FALSE, breaks = ZeroBreaks) +
  geom_qmap() +
  scale_x_longitude() +
  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
  scale_fill_divergent(guide = "none") +
  coord_polar() +
  facet_grid(.~variable, labeller = labeller(variable = label_parsed, lev = lev.lab)) +
  axis_labs_smol +
  tag_facets()
```

@@

```{r}
# indices %>% 
#   copy() %>% 
#   .[lev %in% c(50, 700)] %>% 
#   .[, cumestimate := cumsum(estimate_norm), by = .(lev, term, season(time))] %>% 
#   
#   
#   ggplot(aes(time, cumestimate)) +
#   geom_line(aes(color = term)) +
#   # ggridges::geom_vridgeline(stat="ydensity", scale = -4e8, fill = NA,
#   #                           aes(color = term, width = ..density.., x = min(time - 1e7))) +
#   scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam) +
#   scale_y_continuous(NULL) +
#   # scale_x_datetime(NULL, expand = c(0, 0)) +
#   facet_grid(lev~season(time), labeller = labeller(lev = lev.lab)) +
#   tag_facets()

@@

crosscor <- indices %>% 
  copy() %>% 
  .[, widyr::pairwise_cor(.SD, lev, time, estimate, diag = FALSE), by = .(term)] 
crosscor %>% 
  ggplot(aes(item2, item1)) +
  geom_contour_fill(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
  geom_contour_tanaka(aes(z = correlation), size = 0.1, na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
  geom_text_contour(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1), 
                    color = "white", rotate = FALSE, stroke.color = "black", stroke = 0.2, 
                    skip = 1, size = 2) +
  geom_polygon(data = tri, fill = "white") +
  scale_fill_divergent("Cross-correlation", 
                       breaks = MakeBreaks(binwidth = 0.1),
                       guide = guide_colorstrip_bottom()) +
  scale_x_level(minor_breaks = logbreaks, position = "top") +
  scale_y_level(minor_breaks = logbreaks) +
  coord_equal() +
  facet_grid(. ~ term, labeller = labeller(term = lab_sam)) +
  theme(panel.spacing.x = unit(2, "lines"), strip.placement = "outside") +
  tag_facets(position = "br")
```

@@

```{r compute-regr-pattern}
indices_2d_yearly <- hgt[lev %in% c(main_levs)] %>% 
  indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
  .[, combine_lists(FitLm(hgt_a, sym, asym, se = TRUE),
                    FitLm(hgt_a, full, se = TRUE)),
    by = .(lon, lat, lev)] %>% 
  rm_intercept() %>% 
  .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, lev)] %>% 
  periodic(lon = c(0, 360)) %>% 
  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] 
```

@@

```{r 2d-regr, fig.cap = "Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.env = "figure*", fig.width=7, fig.height=5}
indices_2d_yearly %>% 
  .[lat <= -20] %>% 
  .[lev %in% c(50, 700)] %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), breaks = ZeroBreaks, global.breaks = FALSE) +
  geom_contour_tanaka(aes(z = estimate),
                      breaks = ZeroBreaks,
                      size = 0.3,
                      global.breaks = FALSE) +
  geom_qmap(~.x[lat <= -20]) +
  scale_x_longitude() +
  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
                       limits = c(-85, 85),
                       oob = scales::squish,
                       breaks = MakeBreaks(binwidth = 20, exclude = 0)) +
  scale_linetype(guide = "none") +
  coord_polar() +
  facet_grid(lev ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("rc") +
  axis_labs_smol 
```

Figure \ref{fig:2d-regr} shows the spatial year-long regression #FIXME. Column a are regressinos using the Full SAM, while columns b and c are regression coefficients computed in a multiple regression of geopotential height on the Asymmetric and Symmetric indices at the same time. Thus, they are to be interpreted as the patterns associated with each index, controling for the (linear) effect of the other. (Figure A6 #FIXME illustrates the difference between computing two simple regressions and one multiple regression.)

In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centered in the south pole. The monopole obtained by multiple regression with the Asymmetric and Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel b.1).

In the troposphere, panel a.2 show the well known zonally symmetrical annular mode *contaminated* with zonal asymmetries in the form of a wave-3. The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures. The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is assosiated with an trully annular mode, almost devoid of zonal asymmetries (panel c.2).  #FIXME (agregar algo más?)

```{r wave-amplitude, fig.cap = "Planteray wave amplitude for the regression patterns at 50 and 700 hPa.", fig.width=3, fig.height=4}
indices_2d_yearly %>% 
  .[lat < -20] %>%
  .[lev %in% c(50, 700)] %>% 
  .[, FitWave(estimate, 0:3), by = .(lat, lev, term)] %>% 
  ggplot(aes(lat, amplitude)) +
  geom_line(aes(color = factor(k))) +
  scale_color_brewer("Wave-number", palette = "Dark2") +
  scale_y_continuous("Amplitude") +
  scale_x_latitude(ticks = 20) +
  coord_flip() +
  facet_grid(term~lev, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free_x") +
  tag_facets(position = "tr")
```

The amplitude of each zonal wave number at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean. Comparing between rows, this Figure quantifies the relatively clean separation between the zonally symmetric and zonally asymmetric structures, as its evident how the mixture of waves of the Full field (first row) is very similar to the sum of the waves of the Asymmetric and Symmetric field (second and third row, respectively). The second row of Figure \ref{fig:wave-amplitude} shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel b), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance.

@@

<!-- ```{r compute-regr-pattern-season} -->
<!-- indices_2d <- hgt[lev %in% main_levs] %>%  -->
<!--   indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>%  -->
<!--   .[, rbind(as.data.table(FitLm(hgt_a, sym, asym)), -->
<!--             as.data.table(FitLm(hgt_a, full))), -->
<!--     by = .(lon, lat, lev, season(time))] %>%  -->
<!--   rm_intercept() %>%  -->
<!--   periodic(lon = c(0, 360)) %>%  -->
<!--   .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)]  -->
<!-- ``` -->

<!-- ```{r 2d-regr-700, fig.cap = "Seasonal regression patterns of geopotential height at 700 hPa with the Full, Asymmetric and Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"} -->
<!-- indices_2d %>%  -->
<!--   .[lat <= -20] %>%  -->
<!--   .[lev == 700] %>%  -->
<!--   ggplot(aes(lon, lat)) + -->
<!--   geom_contour_fill(aes(z = estimate),  breaks = AnchorBreaks(0, binwidth = 5, 0), -->
<!--                     global.breaks = FALSE) + -->
<!--   geom_contour_tanaka(aes(z = estimate), -->
<!--                       breaks = AnchorBreaks(0, binwidth = 5, 0), -->
<!--                       size = 0.3, -->
<!--                       global.breaks = FALSE) + -->
<!--   geom_qmap(~.x[lat <= -20]) + -->
<!--   scale_x_longitude() + -->
<!--   scale_y_latitude(limits = c(NA, -20), labels = NULL) + -->
<!--   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), -->
<!--                        limits = c(-55, 55), -->
<!--                        oob = scales::squish, -->
<!--                        breaks = AnchorBreaks(0, binwidth = 5, 0)) + -->
<!--   scale_linetype(guide = "none") + -->
<!--   coord_polar() + -->
<!--   facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) + -->
<!--   tag_facets() + -->
<!--   axis_labs_smol -->
<!-- ``` -->

<!-- ```{r 2d-regr-30, fig.cap = "Seasonal regression patterns of geopotential height at 30 hPa with the Full, Asymmetric and Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"} -->
<!-- indices_2d %>%  -->
<!--   .[lat <= -20] %>%  -->
<!--   .[lev == 30] %>%  -->
<!--   ggplot(aes(lon, lat)) + -->
<!--   geom_contour_fill(aes(z = estimate),  breaks = AnchorBreaks(0, binwidth = 20, 0), -->
<!--                     global.breaks = FALSE) + -->
<!--   geom_contour_tanaka(aes(z = estimate), -->
<!--                       breaks = AnchorBreaks(0, binwidth = 20, 0), -->
<!--                       size = 0.3, -->
<!--                       global.breaks = FALSE) + -->
<!--   geom_qmap(~.x[lat <= -20]) + -->
<!--   scale_x_longitude() + -->
<!--   scale_y_latitude(limits = c(NA, -20), labels = NULL) + -->
<!--   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), -->
<!--                        limits = c(-105, 105), -->
<!--                        oob = scales::squish, -->
<!--                        breaks = AnchorBreaks(0, binwidth = 20, 0)) + -->
<!--   scale_linetype(guide = "none") + -->
<!--   coord_polar() + -->
<!--   facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) + -->
<!--   tag_facets() + -->
<!--   axis_labs_smol -->
<!-- ``` -->

@@

air_regr <- air %>% 
  # .[, air := Anomaly(air), by = .(lon, lat)] %>%
  .[indices_wide[lev == 700], on = .NATURAL] %>% 
  rm_singleton() %>% 
  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
  .[, nas := mean(is.na(air)), by = .(lon, lat, season(time))] %>% 
  .[nas >= 0.15, air := NA] %>% 
  .[, rbind(as.data.table(FitLm(air, asym, sym, se = TRUE)),
            as.data.table(FitLm(air, full, se = TRUE))),
    by = .(lon, lat, season(time))] %>% 
  rm_intercept() 
```

```{r regr-air-season, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas. Gray areas have more than 15\\% of missing data.", fig.width = 10, fig.height = 7, fig.env = "sidewaysfigure"}
air_regr %>% 
  copy() %>% 
  # .[lat <= -61, estimate := NA] %>%
  .[lat >= 65, estimate := NA] %>%
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>% 
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), na.fill = 0,
                    breaks = AnchorBreaks(0, 0.15, 0)) +
  # geom_raster(aes(fill = estimate)) +
  geom_contour2(aes(z = p.value), breaks = 0.05) +
  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 0.5, fill = "gray", color = "gray30") +
  geom_qmap(~.x[lat <= 0]) +
  scale_x_longitude() +
  scale_y_latitude(labels = NULL, limits = c(NA, 0)) +
  scale_fill_divergent("Surface temperature",  limits = c(-0.8, .8), oob = scales::squish,
                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.15, 0)) +
  coord_polar() +
  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr", tag_suffix = "") +
  axis_labs_smol
```

@@

??

>>>>

---
twocol: FALSE  #  Use FALSE for submission. 
title: Title here
author1: 
  name: Elio Campitelli
  email: elio.campitelli@cima.fcen.uba.ar
author2: 
  name: Leandro Díaz
affiliation: "CIMA UBA blablabla"
extra_authors: 
  - name: Carolina Vera
abstract: |
  Enter the text of your abstract here. This is a sample American Meteorological Society (AMS) \LaTeX\ template.  This document provides authors with instructions on the use of the AMS \LaTeX\ template.  Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document. The template.tex file should be modified by authors for their own manuscript.
statement: |
  This is significant becasue I wrote it.
bibliography: AsymSAM.bib
output: 
  rticles::ams_article:
    latex_engine: "xelatex"
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
---

```{r setup, echo = FALSE,warning=FALSE, message = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.path = "",
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.extra = 41,
  cache.path = "../cache/",
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  out.extra = ""
)
library(metR)
library(data.table)
library(ggplot2)
library(magrittr)
library(ggperiodic)
library(asymsam)
library(latex2exp)
library(tagger)

theme_set(theme_asymsam(base_size = 10))
ZeroBreaks <- AnchorBreaks(0, NULL, 0)
axis_labs_smol <- theme(axis.text = element_text(size = 6))
lev.lab <- AddSuffix(" hPa")
no_grid <- theme(panel.grid = element_blank())

@@

combine_lists <- function(...) {
  # browser(expr = length(x) != length(y))
  X <- list(...)
  names <- names(X[[1]])
  names(names) <- names
  lapply(names, function(n) {
    unlist(lapply(X, function(x) x[[n]]))
  })
}

lab_sam <-  c(full = "Full", 
              asym = "Asymmetric",
              sym  = "Symmetric")
```

@@

```{r compute-SAM}
SAM <- hgt[lat <= -20] %>% 
  .[lev == 700] %>% 
  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
    by = .(lev)]

@@

We used monthly geopotential height at 2.5 longitude by 2.5 latitude resolution from ERA5 [@hersbach] for the period 1979 to 2018.

@@

```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height. Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right). Arbitrary units.", fig.width=6, fig.height=2.5, fig.env="figure*"}
sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 
  setnames("hgt", "full") %>% 
  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev)]

@@

sam_plot <- sam_sep %>% 
  .[lev %in% c(700)] %>%
  rm_singleton() %>% 
  melt(id.vars = c("lon", "lat")) %>% 
  .[, value := value + rnorm(.N, sd = 1e-6)] %>% 
  .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)]

@@

sam_plot %>% 
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +
  # geom_contour(aes(z = value, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
  geom_contour_fill(aes(z = value), global.breaks = FALSE, breaks = ZeroBreaks) +
  geom_contour_tanaka(aes(z = value), global.breaks = FALSE, breaks = ZeroBreaks) +
  geom_qmap() +
  scale_x_longitude() +
  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
  scale_fill_divergent(guide = "none", high = "#731C1F", low = "#323071") +
  scale_fill_divergent(guide = "none") +
  coord_polar() +
  facet_grid(.~variable, labeller = labeller(variable = label_parsed, lev = lev.lab)) +
  axis_labs_smol +
  no_grid +
  tag_facets()
```

@@

```{r test-aao}
# test that my indices hace the correct sign. The Full index needs to be **positively** correlated
# with the AAO index.
aao <- rsoi::download_aao(TRUE, data_path("raw", "aao.csv")) %>% 
  as.data.table() %>% 
  .[, .(time = lubridate::as_datetime(Date), aao = AAO)]

cor_aao <- indices[term == "full" & lev == 700] %>% 
  .[aao, on = .NATURAL] %>% 
  na.omit() %>% 
  .[, cor(estimate, aao)]

if (sign(cor_aao) != 1) {
  stop("AAO and full index are not positively correlated!")
}
```

@@

crosscor <- indices %>% 
  copy() %>% 
  .[, widyr::pairwise_cor(.SD, lev, time, estimate, diag = TRUE), by = .(term)] 
crosscor %>% 
  ggplot(aes(item2, item1)) +
  geom_contour_fill(aes(z = correlation), breaks = MakeBreaks(binwidth = 0.1)) +
  geom_contour_tanaka(aes(z = correlation), size = 0.1, na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
  geom_text_contour(aes(z = correlation), breaks = MakeBreaks(binwidth = 0.1), 
                    color = "white", rotate = FALSE, stroke.color = "black", stroke = 0.2, 
                    skip = 1, size = 2) +
  geom_polygon(data = tri, fill = "white") +
  scale_fill_divergent("Cross-correlation", 
                       breaks = MakeBreaks(binwidth = 0.1),
                       guide = guide_colorstrip_bottom()) +
  scale_x_level(minor_breaks = logbreaks, position = "top") +
  scale_y_level(minor_breaks = logbreaks) +
  coord_equal() +
  facet_grid(. ~ term, labeller = labeller(term = lab_sam)) +
  theme(panel.spacing.x = unit(2, "lines"), strip.placement = "outside") +
  tag_facets(position = "br")
```

@@

```{r compute-regr-pattern}
indices_2d_yearly <- hgt[lev %in% c(main_levs)] %>% 
  indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
  .[, combine_lists(FitLm(hgt_a, sym, asym, se = TRUE),
                    FitLm(hgt_a, full, se = TRUE)),
    by = .(lon, lat, lev)] %>% 
  rm_intercept() %>% 
  .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, lev)] %>% 
  periodic(lon = c(0, 360)) %>% 
  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>% 
  .[, season := "Annual"]
```

@@

```{r 2d-regr, fig.cap = "Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.env = "figure*", fig.width=7, fig.height=5}
indices_2d_yearly %>% 
  .[lat <= -20] %>% 
  .[lev %in% c(50, 700)] %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, exclude =  0), global.breaks = FALSE) +
  geom_contour_tanaka(aes(z = estimate),
                      breaks = AnchorBreaks(0, exclude =  0),
                      size = 0.3,
                      global.breaks = FALSE) +
  geom_qmap(~.x[lat <= -20]) +
  scale_x_longitude() +
  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
                       limits = c(-45, 45),
                       oob = scales::squish,
                       breaks = AnchorBreaks(0, 10, 0)) +
  scale_linetype(guide = "none") +
  coord_polar() +
  facet_grid(lev ~ term) + # labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("rc") +
  axis_labs_smol  +
  no_grid
```

Figure \ref{fig:2d-regr} shows the spatial year-long regression #FIXME. Column a are regressinos using the Full SAM, while columns 2 and 3 are regression coefficients computed in a multiple regression of geopotential height on the Asymmetric and Symmetric indices at the same time. Thus, they are to be interpreted as the patterns associated with each index, controling for the (linear) effect of the other. (Figure A6 #FIXME illustrates the difference between computing two simple regressions and one multiple regression.)

In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centered in the south pole. The monopole obtained by multiple regression with the Asymmetric and Symmetric SAM (panel a.3) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel a.2).

In the troposphere, panel b.1 shows the well known zonally symmetrical annular mode *contaminated* with zonal asymmetries in the form of a wave-3. The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures. The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is assosiated with an trully annular mode, almost devoid of zonal asymmetries (panel b.3). note that the wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern and asociated with @raphael2004's ZW3 index (see Figure 1 from that paper). #FIXME (agregar algo más?)

```{r calculate-raphael}
# 50°E, 166°E and 76°W. 
raphael <- ReadNetCDF(ERA5(), 
           subset = list(time = c("1979-01-01", "2018-12-31"),
                         latitude = -49,
                         level = 500),
           vars = c(hgt = "z")) %>% 
  normalise_coords() %>% 
  .[, hgt := hgt/9.8] %>% 
  .[lon %~% ConvertLongitude(c(-50, -166, 76))] %>% 
  .[, hgt_s := frollmean(hgt, 3, NA, align = "center"), by = .(lon, lat)] %>% 
  na.omit() %>% 
  .[, hgt_s := scale(hgt_s), by = .(lon, lat, month(time))] %>% 
  .[, .(raphael = mean(hgt_s)), by = time]
```

```{r wave-amplitude, fig.cap = "Planteray wave amplitude for the regression patterns at 50 and 700 hPa. Note the varying x axis.", fig.width=3, fig.height=3}
indices_2d_yearly %>% 
  .[lat < -20] %>%
  .[lev %in% c(50, 700)] %>% 
  .[, FitWave(estimate, 0:3), by = .(lat, lev, term)] %>% 
  ggplot(aes(lat, amplitude)) +
  geom_line(aes(color = factor(k))) +
  scale_color_brewer("Wave-number", palette = "Dark2") +
  scale_y_continuous("Amplitude") +
  scale_x_latitude(ticks = 20) +
  coord_flip() +
  facet_grid(lev~term, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free") +
  tag_facets("cr", position = "tr")  +
  no_grid +
  theme(strip.text = element_text(size = 7))
```

The amplitude of each zonal wave number at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean. Comparing between rows, this Figure quantifies the relatively clean separation between the zonally symmetric and zonally asymmetric structures, as its evident how the mixture of waves of the Full field (column a) is very similar to the sum of the waves of the Asymmetric and Symmetric field (columns b and c, respectively). Column b of Figure \ref{fig:wave-amplitude} shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel b.1), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance (panel b.2).

@@

```{r enso-cor}
mei_sam_cor <- rsoi::download_mei(TRUE, data_path("derived", "mei.csv")) %>% 
  as.data.table() %>% 
  .[, .(time = lubridate::as_datetime(Date), mei = MEI)] %>% 
  .[indices, on = "time"] %>% 
  .[, with(cor.test(estimate, mei), 
           list(cor = estimate, 
                low = conf.int[1], 
                upp = conf.int[2])), by = .(lev, term)] 
# enso_sam_cor %>% 
#   ggplot(aes(lev, cor)) +
#   geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha = 0.2) +
#   geom_line(aes(color = term)) +
#   scale_x_level() +
#   coord_flip()

cor_texts <- mei_sam_cor[lev == 700, 
                          paste0(signif(cor, 2)), # " (CI: ", signif(low, 2), "–", signif(upp, 2), ")"), 
                          by = term] %>% 
  dcast(1~term)

mei_sam_partial <- rsoi::download_mei(TRUE, data_path("derived", "mei.csv")) %>% 
  as.data.table() %>% 
  .[, .(time = lubridate::as_datetime(Date), mei = MEI)] %>% 
  .[indices_wide, on = "time"] %>% 
  .[lev == 700] %>% 
  .[, partial_cor(mei, asym, sym)] %>% 
  .[, partial_correlation := signif(partial_correlation, 2)] %>% 
  dcast(.~term)
```

The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the teleconnection pattern associated with the ENSO. Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO. The correlation between the full SAM and the ENSO as measured by the Multivariate Enso Index [@wolter2011] is `r cor_texts$full`. This relationship is captured entirely the Asymmetric SAM, as this index has a partial correlation of `r mei_sam_partial$asym` with the MEI, whereas the Symmetric SAM has null partial correlation with the MEI.

@@

# mei <- rsoi::download_mei() %>% 
#   as.data.table() %>% 
#   .[, .(time = lubridate::as_datetime(Date), mei = MEI)]

air_regr <- air %>% 
  # .[, air := Anomaly(air), by = .(lon, lat)] %>%
  .[indices_wide[lev == 700], on = .NATURAL] %>% 
  # mei[., on = .NATURAL] %>% 
  rm_singleton() %>% 
  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
  .[, nas := mean(is.na(air)), by = .(lon, lat, season(time))] %>% 
  .[nas >= 0.15, air := NA] %>% 
  .[, rbind(as.data.table(FitLm(air, asym, sym, se = TRUE)),
            as.data.table(FitLm(air, full, se = TRUE))),
    by = .(lon, lat, season(time))] %>% 
  rm_intercept() 
```

```{r regr-air-season, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas. Gray areas have more than 15\\% of missing data.", fig.width = 6, fig.height = 8, fig.env = "figure*"}
air_regr %>% 
  # .[season == "DJF"] %>% 
  copy() %>% 
  # .[lat <= -61, estimate := NA] %>%
  .[lat >= 65, estimate := NA] %>%
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
  # .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>%
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), na.fill = 0,
                    breaks = AnchorBreaks(0, 0.15, 0)) +
  geom_contour_tanaka(aes(z = estimate), na.fill = 0,
                      breaks = AnchorBreaks(0, 0.15, 0)) +
  # geom_raster(aes(fill = estimate)) +
  geom_contour_fill(aes(z = p.value), breaks = c(0, 0.05), color = "black", fill = "gray50", alpha = 0.2) +
  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), fill = "gray", color = "gray30") +
  geom_qmap(~.x[lat <= 0]) +
  scale_x_longitude() +
  scale_y_latitude(labels = NULL, limits = c(NA, 0)) +
  scale_fill_divergent("Surface temperature",  limits = c(-0.8, .8), oob = scales::squish,
                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.15, 0)) +
  coord_polar() +
  facet_grid(season ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr", tag_suffix = "") +
  no_grid +
  axis_labs_smol
```

@@

Regression of the SAM indicies with seasonal mean precipitation are shown Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australia and New Zealand, and South America respectively. (We didn't detect any significant signal in South Africa.)

```{r}
indices_2d_seasonal <- hgt[lev %in% 700] %>% 
  indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>% 
  .[, combine_lists(FitLm(hgt_a, sym, asym, se = TRUE),
                    FitLm(hgt_a, full, se = TRUE)),
    by = .(lon, lat, season(time))] %>% 
  rm_intercept() %>% 
  .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
  periodic(lon = c(0, 360)) %>% 
  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)]

indices_2d <- rbind(indices_2d_yearly[lev == 700] %>% .[, lev := NULL], indices_2d_seasonal) %>% 
  .[, season := forcats::fct_relevel(season, "Annual")] 
```
